Region-Network Hierarchical Sparsity Priors for\\ High-Dimensional Inference in Brain Imaging

### 2016-02-03 paper/icml2016/example_paper.tex
ICMLize II

\twocolumn[
\icmltitle{Region-Network Hierarchical Sparsity Priors for\\
High-Dimensional Inference in Brain Imaging}

@@

\paragraph{Implementation.}
The analyses were performed in Python.
We used \textit{nilearn} to handle
the large quantities of neuroimaging data 
\cite{abrah14}
and
\textit{Theano} for automatic, numerically stable
differentiation of symbolic computation graphs
\cite{bastien2012theano, bergstra2010theano}.
All Python scripts that generated the results are
accessible online for reproducibility and reuse
% (\url{http://github.com/anonymous/anonymous}).
(\url{http://github.com/banilo/nips2015}).

all algorithm from a same software library -> SPAMs

%
\paragraph{Data.}
As the currently biggest openly-accessible reference dataset,
we chose resources from the Human Connectome Project (HCP)
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500
healthy HCP participants (cf. Appendix for details on datasets).
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants (Table \ref{table_tasks}).
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common $60\times72\times60$ space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
The supervised analyses were thus based on labeled HCP task maps with
79,941 voxels of interest representing z-values in gray matter.

\begin{table*}[h]
  \resizebox{0.98\textwidth}{!}{%
  \begin{tabular}{l|l|l}
    \hline
  {\bf Cognitive Task} & {\bf Stimuli}                         & {\bf Instruction for participants}                                                \\ \hline
  1 Reward             & \multirow{2}{*}{Card game}            & \multirow{2}{*}{Guess the number of a mystery card for gain/loss of money}        \\ \cline{1-1}
  2 Punish             &                                       &                                                                                   \\ \hline
  3 Shapes             & Shape pictures                        & Decide which of two shapes matches another shape geometrically                    \\ \hline
  4 Faces              & Face pictures                         & Decide which of two faces matches another face emotionally                        \\ \hline
  5 Random             & \multirow{2}{*}{Videos with objects}  & \multirow{2}{*}{Decide whether the objects act randomly or intentionally} \\ \cline{1-1}
  6 Theory of mind     &                                       &                                                                                   \\ \hline
  7 Mathematics        & Spoken numbers                        & Complete addition and subtraction problems                                        \\ \hline
  8 Language           & Auditory stories                      & Choose answer about the topic of the story                                        \\ \hline
  9 Tongue movement    & \multirow{3}{*}{Visual cues}          & Move tongue                                                                       \\ \cline{1-1} \cline{3-3} 
  10 Food movement     &                                       & Squeezing of the left or right toe                                                \\ \cline{1-1} \cline{3-3} 
  11 Hand movement     &                                       & Tapping of the left or right finger                                               \\ \hline
  12 Matching          & \multirow{2}{*}{Shapes with textures} & Decide whether two objects match in shape or texture                             \\ \cline{1-1} \cline{3-3} 
  13 Relations         &                                       & Decide whether object pairs differ both along either shape or texture             \\ \hline
  14 View Bodies       & Pictures                              & Passive watching                                                                   \\ \hline
  15 View Faces        & Pictures                              & Passive watching                                                                   \\ \hline
  16 View Places       & Pictures                              & Passive watching                                                                   \\ \hline
  17 View Tools        & Pictures                              & Passive watching                                                                   \\ \hline
  18 Two-Back          & Various pictures                      & Indicate whether current stimulus is the same as two items earlier                \\ \hline
  \end{tabular}
}
\vspace{-0.2cm}
\caption{\textbf{Description of psychological tasks to predict.}}
\label{table_tasks}
\end{table*}

These labeled data were complemented by unlabeled activity maps
from HCP acquisitions of unconstrained resting-state activity
\cite{smith2013resting}.
These reflect brain activity in the absence of controlled thought.
In sum, the HCP rest data concatenated
8000 unlabeled, noise-cleaned rest maps with
40 brain maps from each of 200 randomly selected participants.

We were further interested in the utility of the
optimized low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classification problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
Analogous to HCP data, the second task dataset thus incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

sparse statistical models have only few nonzero parameters

@@

\begin{figure*}
\begin{centering}
\includegraphics[width=1.00\textwidth]{../figures/reg_net_prior_colin.pdf}
\end{centering}
\vspace{-0.6cm}
\caption{\textbf{Building blocks of the region-network tree.}
Depicts neurobiological priors introduced into the classification problem 
by hierarchical structured sparsity.
\textit{Left:} Continuous, partially overlapping brain network priors
(\textit{hot-colored}, taken from \cite{smith2009})
accommodate the functional integration
perspective of brain organization.
\textit{Right:} Discrete, non-overlapping brain region priors
(\textit{single-colored}, taken from \cite{crad12})
accommodate the functional segregation perspective.
\textit{Middle:} These two types of predefined voxel groups are incorporated
into hierarchical priors of parent networks with their
descending region nodes.
\textit{Top to bottom:} Four examplary region-network priors
are shown, including
the early cortex that processes
visual and sound information from the environment,
a well-known attentional circuit in the left brain hemisphere,
and
the cerebellum that is involved in motor behavior.
}
\label{fig_priors}
\end{figure*}

\begin{figure*}
\begin{centering}
\includegraphics[width=1.00\textwidth]{../figures/sparsities.pdf}
\end{centering}
\vspace{-0.6cm}
\caption{\textbf{Model performance across sparsity priors.}
Compares the performance of logistic regression estimators
with 4 different structured and unstructered sparsity terms
in classifying neural activity from 18 psychological tasks.
The class-wise precision and recall metrics
were obtained on the same test set.
%
Unstructured $\ell_1$-penalized logistic regression (\textit{upper left})
imposed a minimum of relevant brain voxels without
assuming special structure.
Structured $\ell_1/\ell_2$-block norm with additional $\ell_1$ term
(\textit{lower left})
imposed region compartments, but naive to network structure.
Structured trace-norm penalization (\textit{lower right})
imposed low-rank structure
with sparsity of network patterns, but naive to region structure.
Structured $\ell_1/\ell_2$-block-norm with a hierarchy of
both region and network priors (\textit{upper right})
exhibited the best out-of-sample performance.
%
A priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting psychological tasks from
brain maps.
}
\label{fig_sparsities}
\end{figure*}

@@

\begin{figure*}
\begin{centering}
\includegraphics[width=0.80\textwidth]{../figures/reg_net_ratio_colin.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Support recovery as a function of
region and network emphasis.}
The relative impact of the region and network priors
on model selection
is systematically varied against each other.
This region-network ratio (\textit{upper fractions}) weighted voxel groups
to priviledge sparse models in function space
that acknowledge known brain region neighborhoods
(\textit{left columns}) or
known brain networks neighborhoods
(\textit{right columns}).
Among the 18 classes, the model weights are shown for the psychological
tasks (\textit{from top to bottom}): tongue movement, listening stories,
taking somebody else's perspective (ToM, "theory of mind"),
as well as
viewing locations and tools.
The 18-class out-of-sample accuracy is indicated
on the \textit{bottom} and
the class-wise mean neural activty
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
%
Different emphasis on regions versus networks
in hierarchical structured sparsity can
yield comparable model performance.
%
Favoring region versus network structure during model selection
recovers complementary aspects of the neural activity pattern.
%
Equal region and network emphasis yields more dispersed,
less interpretable predictive model choices.
}
\label{fig_regnetratio}
\end{centering}
\end{figure*}

\paragraph{Support recovery as a function of region and network emphasis.}
Finally, the relative importance of the
region and network priors within the hierarchical tree prior
was quantified (Figure \ref{fig_regnetratio}).
The group weight $\eta_g$ of region priors was multiplied with a
region-network ratio, while the
group weight $\eta_g$ of network priors was divided by that
region-network ratio. For instance, a region-network ratio of 3
increased the relative importance of known region structure
by multiplying $\frac{3}{1}$ to
$\eta_g$ of all region group penalties
and multiplying
$\frac{1}{3}$ to $\eta_g$ of all network group penalties.
The data splitting cross-validation scheme was identical to the
above modelling experiments.

@@

\section{Discussion}
Relevant structure in neuroimaging data has long been investigated
according to two separate organizational principles:
functional segregation into discrete brain regions
\cite{passingham2002}
and functional integration by interregional brain networks
\cite{sporns14nn}.
%
This proof-of-concept study demonstrates the simultaneous exploitation of
both these neurobiological compartments
for sparse variable selection and high-dimensional prediction
in a reference dataset.
%
Introducing existing domain knowledge into model selection
allowed privileging members of the function space
that are most neurobiologically plausible.
%
Domain-informed hierarchical structured sparsity is shown to enhance
both model interpretability and generalization performance,
although these statistical-learning goals are typically in conflict.

@@

\small
% \bibliographystyle{plainnat}
\bibliographystyle{icml2016}
\bibliography{paper_refs}
% \bibliography{paper_refs}

>>>>

\twocolumn[
\icmltitle{Region-Network Hierarchical Sparsity Priors for\\
High-Dimensional Inference in Brain Imaging}
% \icmltitle{Region-Network Hierarchical Sparsity Priors for\\
% High-Dimensional Inference in Brain Imaging}

@@

\paragraph{Implementation.}
All analyses were performed in Python.
We used \textit{nilearn} to process and resphape
the extensive neuroimaging data 
\cite{abrah14},
\textit{scikit learn} to design machine-learning
data processing pipelines
\cite{pedr11},
and
\textit{SPAMs} for numerically optimized
implementations of the sparse estimators
(\url{http://spams-devel.gforge.inria.fr/}).
All Python scripts that generated the results are
accessible online for reproducibility and reuse
(\url{http://github.com/anonymous/anonymous}).
% (\url{http://github.com/banilo/nips2015}).

%
\paragraph{Data.}
As the currently biggest openly-accessible reference dataset,
we chose resources from the Human Connectome Project (HCP)
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500
healthy HCP participants (cf. Appendix for details on datasets).
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants.
% across participants (Table \ref{table_tasks}).
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common $60\times72\times60$ space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
The supervised analyses were thus based on labeled HCP task maps with
79,941 voxels of interest representing z-values in gray matter.

% \begin{table*}[h]
%   \resizebox{0.98\textwidth}{!}{%
%   \begin{tabular}{l|l|l}
%     \hline
%   {\bf Cognitive Task} & {\bf Stimuli}                         & {\bf Instruction for participants}                                                \\ \hline
%   1 Reward             & \multirow{2}{*}{Card game}            & \multirow{2}{*}{Guess the number of a mystery card for gain/loss of money}        \\ \cline{1-1}
%   2 Punish             &                                       &                                                                                   \\ \hline
%   3 Shapes             & Shape pictures                        & Decide which of two shapes matches another shape geometrically                    \\ \hline
%   4 Faces              & Face pictures                         & Decide which of two faces matches another face emotionally                        \\ \hline
%   5 Random             & \multirow{2}{*}{Videos with objects}  & \multirow{2}{*}{Decide whether the objects act randomly or intentionally} \\ \cline{1-1}
%   6 Theory of mind     &                                       &                                                                                   \\ \hline
%   7 Mathematics        & Spoken numbers                        & Complete addition and subtraction problems                                        \\ \hline
%   8 Language           & Auditory stories                      & Choose answer about the topic of the story                                        \\ \hline
%   9 Tongue movement    & \multirow{3}{*}{Visual cues}          & Move tongue                                                                       \\ \cline{1-1} \cline{3-3} 
%   10 Food movement     &                                       & Squeezing of the left or right toe                                                \\ \cline{1-1} \cline{3-3} 
%   11 Hand movement     &                                       & Tapping of the left or right finger                                               \\ \hline
%   12 Matching          & \multirow{2}{*}{Shapes with textures} & Decide whether two objects match in shape or texture                             \\ \cline{1-1} \cline{3-3} 
%   13 Relations         &                                       & Decide whether object pairs differ both along either shape or texture             \\ \hline
%   14 View Bodies       & Pictures                              & Passive watching                                                                   \\ \hline
%   15 View Faces        & Pictures                              & Passive watching                                                                   \\ \hline
%   16 View Places       & Pictures                              & Passive watching                                                                   \\ \hline
%   17 View Tools        & Pictures                              & Passive watching                                                                   \\ \hline
%   18 Two-Back          & Various pictures                      & Indicate whether current stimulus is the same as two items earlier                \\ \hline
%   \end{tabular}
% }
% \vspace{-0.2cm}
% \caption{\textbf{Description of psychological tasks to predict.}}
% \label{table_tasks}
% \end{table*}

@@

\begin{figure*}
\begin{centering}
\includegraphics[width=0.8\textwidth]{../figures/reg_net_prior_colin.pdf}
\end{centering}
\vspace{-0.2cm}
\caption{\textbf{Building blocks of the region-network tree.}
Depicts neurobiological priors introduced into the classification problem 
by hierarchical structured sparsity.
\textit{Left:} Continuous, partially overlapping brain network priors
(\textit{hot-colored}, taken from \cite{smith2009})
accommodate the functional integration
perspective of brain organization.
\textit{Right:} Discrete, non-overlapping brain region priors
(\textit{single-colored}, taken from \cite{crad12})
accommodate the functional segregation perspective.
\textit{Middle:} These two types of predefined voxel groups are incorporated
into hierarchical priors of parent networks with their
descending region nodes.
\textit{Top to bottom:} Four examplary region-network priors
are shown, including
the early cortex that processes
visual and sound information from the environment,
a well-known attentional circuit in the left brain hemisphere,
and
the cerebellum that is involved in motor behavior.
}
\label{fig_priors}
\end{figure*}

\begin{figure*}
\begin{centering}
% \includegraphics[width=1.00\textwidth]{../figures/sparsities.pdf}
\includegraphics[width=0.55\textwidth]{../figures/ROC_ALL_perclass.png}
\end{centering}
\vspace{-0.4cm}
\caption{\textbf{Model performance across sparsity priors.}
Compares the performance of logistic regression estimators
with 4 different structured and unstructered sparsity terms
in classifying neural activity from 18 psychological tasks.
The class-wise area under the curve (AUC)
is obtained on the same test set.
%
\textit{Hierarchical Tree Sparsity}: Structured
$\ell_1/\ell_2$-block-norm with a hierarchy of
both region and network priors
exhibited the best out-of-sample performance.
\textit{Lasso}: Unstructured $\ell_1$-penalized logistic regression
imposed a minimum of relevant brain voxels without
assuming special structure.
\textit{Elastic-Net}: Unstructured logistic regression
with interpolation between $\ell_1$- and $\ell_2$-norm 
imposed an equilibrium between sparsity and model fit.
\textit{Trace-norm}: Structured trace-norm penalization
imposed low-rank structure
with sparsity of network patterns, but naive to region structure.
\textit{(Sparse) Group Sparsity}: Structured $\ell_1/\ell_2$-block
norm with additional $\ell_1$ term
imposed region compartments, but naive to network structure.
%
A priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting psychological tasks from
brain maps.
}
\label{fig_sparsities}
\end{figure*}

@@

\begin{figure*}
\begin{centering}
\includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Support recovery as a function of
region and network emphasis.}
The relative impact of the region and network priors
on model selection
is systematically varied against each other.
This region-network ratio (\textit{upper fractions}) weighted voxel groups
to priviledge sparse models in function space
that acknowledge known brain region neighborhoods
(\textit{left columns}) or
known brain networks neighborhoods
(\textit{right columns}).
Among the 18 classes, the model weights are shown for the psychological
tasks (\textit{from top to bottom}): tongue movement, listening stories,
taking somebody else's perspective (ToM, "theory of mind"),
as well as
viewing locations and tools.
The 18-class out-of-sample accuracy is indicated
on the \textit{bottom} and
the class-wise mean neural activty
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
%
Different emphasis on regions versus networks
in hierarchical structured sparsity can
yield comparable model performance.
%
Favoring region versus network structure during model selection
recovers complementary aspects of the neural activity pattern.
%
Equal region and network emphasis yields more dispersed,
less interpretable predictive model choices.
}
\label{fig_regnetratio}
\end{centering}
\end{figure*}

\paragraph{Support recovery as a function of region-network emphasis.}
Finally, the relative importance of the
region and network priors within the hierarchical tree prior
was quantified (Figure \ref{fig_regnetratio}).
The group weight $\eta_g$ of region priors was multiplied with a
region-network ratio, while the
group weight $\eta_g$ of network priors was divided by that
region-network ratio. For instance, a region-network ratio of 3
increased the relative importance of known region structure
by multiplying $\frac{3}{1}$ to
$\eta_g$ of all region group penalties
and multiplying
$\frac{1}{3}$ to $\eta_g$ of all network group penalties
(Table \ref{table_reg_net_ratio}).

\begin{table*}[]
\centering
\caption{18-class out-of-sample performance by region-network emphasis}
\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
\textbf{Reg-Net Ratio} & 500 & 100  & 50   & 10   & 5    & 2    & 1    & $\frac{1}{2}$  & $\frac{1}{5}$  & $\frac{1}{10}$ & $\frac{1}{50}$ & $\frac{1}{100}$ & $\frac{1}{250}$ & $\frac{1}{500}$ & $\frac{1}{1000}$ \\ \hline
\textbf{Accuracy {[}\%{]}}      & 3.8          & 41.7 & 63.7 & 89.9 & 89.9 & 88.5 & 87.9 & 87.9 & 87.9 & 87.8 & 88.6 & 87.8  & 85.6  & 67.2  & 39,3  
\end{tabular}
\label{table_reg_net_ratio}
\end{table*}

@@

\section{Conclusion}
Relevant structure in neuroimaging data has long been investigated
according to two separate organizational principles:
functional segregation into discrete brain regions
\cite{passingham2002}
and functional integration by interregional brain networks
\cite{sporns14nn}.
%
This proof-of-concept study demonstrates the simultaneous exploitation of
both these neurobiological compartments
for sparse variable selection and high-dimensional prediction
in a reference dataset.
%
Introducing existing domain knowledge into model selection
allowed privileging members of the function space
that are most neurobiologically plausible.
%
Domain-informed hierarchical structured sparsity is shown to enhance
both model interpretability and generalization performance,
although these statistical-learning goals are typically in conflict.

@@

\small
% \bibliographystyle{plainnat}
\bibliographystyle{icml2016}
\bibliography{../paper_refs}
% \bibliography{paper_refs}
